Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2022-01-17 analysis/paper/London_HauloutBeahvior.Rmd
CHANGE FONT TO Times (via newtxtext)
ADDRESS LORI Q FINAL COMMENTS
lori's comments were largely related to references. still need to address improvements to description of weather covariate response as in #26

@@

  - \usepackage {newpxtext}
  - \usepackage {newpxmath}

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding, and molting [@boveng2009;
@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
complicate predictions about the ultimate effects of changes in sea ice on the
behavior, health, abundance, and distribution of these seals. To date, indices
of seal health sampled during periods of declining sea ice differ regionally
[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
timing of reproductive and molting behavior is generally lacking, so it is
difficult to predict the readiness with which ice-associated seal species will
be able to adapt to future changes (e.g., by adjusting pupping or molting
schedules to earlier dates or different locales). This is further complicated by
the spatio-temporal variation in the phenology of these life history events
within regions and throughout their full ranges. Additionally, trends in
abundance from large scale surveys of these species are unknown, so it is
difficult to assess the effect, if any, declines in sea ice habitat have had, or
will have, on seal densities.

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a resting platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
sea-ice but also within snow lairs built during the spring. The remoteness of
the Bering and Chukchi seas mean direct scientific observation of seal behavior
is impractical. Thus, bio-logging devices are especially useful tools for
collecting key information on movement and haul-out behavior for these species.

Bio-logging records of time spent out of the water provide valuable data for
identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly et al. [-@kelly2010] using VHF radio tags and
direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@reder2003; @udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

```{r getdata}
targets::tar_load(analysis_data, store = here::here("_targets"))
targets::tar_load(grid, store = here::here("_targets"))
targets::tar_load(deploy_table, store = here::here("_targets"))
```
Data used in our study came from bio-loggers deployed on bearded, ribbon, and
spotted seals in the Bering, Chukchi, and Beaufort seas by multiple
organizations as part of collaborative investigations from 2005 through 2020.
Seals were captured using nets and bio-loggers attached during ship-based
capture trips and from coastal Alaska communities. Ship-based capture events
occurred during spring near the southern ice edge in the Bering Sea. Land-based
capture events occurred from May to October, generally between the coastal
communities of Scammon Bay, Alaska, in the Bering Sea, and Utqiaġvik, Alaska,
near the junction of the Chukchi and Beaufort seas, from 2005 to 2020. We refer
readers to the primary literature for detailed capture and bio-logger attachment
methods (Supplemental Material, Table).

Haul-out behavior data from bio-loggers deployed on 
`r dplyr::n_distinct(deploy_table$speno)` bearded, ribbon, and spotted seals were
subset to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is at the surface and can slightly
bias the hourly percent-dry values reported by the bio-logger. For this study,
in cases where both bio-logger types were deployed, hourly percent dry
observations from the flipper tag were preferred.

Field identification of age class can be inexact, particularly when discerning
subadults from adults. Sex and age class (non-dependent _young-of-the-year_,
sexually immature _subadults_, and mature _adults_) were estimated at the time
of deployment by various combinations of length, claw growth ridges [@burns1969;
@burns1983; @mclaren1958a], and pelage characteristics for some species. In the
case of ribbon seals, subadults often have less distinct ribbons than adults.
Bearded seal subadults will often have a spotted pattern in the pelage that is
not seen in adults. Spotted seal pelage cannot be used to reliably discern
adults from subadults. Despite these challenges, we feel the age classifications
used in this analysis are useful in testing for age-related effects on haul-out
behavior. Seals determined to be less than one year were classified as
young-of-the-year. Subadults are those seals likely greater than one year of age
but not yet sexually mature. Adults are likely sexually mature and older than
four years. For those bio-loggers deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on 1 March of the
following year. Table \@ref(tab:tblDeploy) provides a summary of these
deployments and data received from them.

@@

Haul-out behavior data were recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and a percent of the
hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
bio-logger, hourly percent-dry data were rounded to the nearest 10 percent
inclusive of 0 and 100 percent and additional values at 3 and 98 percent. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Bio-loggers were, generally, deployed and
programmed in a manner to maximize data transmission during the spring pupping
and molting period; however, hourly percent-dry data were not always
successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (i.e. tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
extra-terrestrial atmospheric interference. Fortunately, missing records do not
seem to greatly bias inference about haul-out probabilities [@conn2012b].

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployment on a ribbon seal over two years during the months of March, April, and May. Areas of the plot indicated with an 'X' represent missing data that were not successfully transmitted from the tag."}

@@

p <- p + facet_grid(year~month, scales = "free_x")
p <- p + scale_x_continuous(breaks = c(5,15,25), expand = c(0,0)) +
  scale_y_continuous(breaks = c(4,12,20))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  theme(panel.spacing = unit(0,"line")) +
  ggtitle("Observation records are incomplete because not all data are transmitted",
          subtitle = "percent-dry data for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)") 
p <- p + 
  geom_point(aes(x = day, y = hour), data = missing_hours, 
             color = "black", shape = 4, size=0.8)
p
```

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and animal location. We explored the use
of a continuous-time correlated random walk [@devins.johnson2008] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50m. Any days where haul-out observations were
present but location data were missing we used the last calculated weighted
average daily location, and any days where the location intersected with land
were removed from the data set. We recognized that bearded and spotted seals
haul out on land. However, assessing the relationship between haul-out behavior
and weather covariates and their availability for aerial surveys on land was
outside the scope of this study. Additionally, any daily locations on land are
likely more reflective of coordinate averaging and measurement error instead of
actual use of coastal features. Figure \@ref(fig:dataMap) shows the spatial
distribution weighted locations with available haul-out behavior data for each
species across the study area.

@@

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill
(represented by a *wind:temperature* interaction [@udevitz2009]), and
day-of-year and time-of-day (solar hour) interactions to permit diurnal patterns
to change from March to July. Ribbon and spotted seal models included age-sex
class and interactions between day-of-year and age-sex class, but we omitted
these from bearded seal models due to poor representation of age-sex classes
(Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial
range and we were interested in possible differences in the timing of haul-out
behavior along a latitudinal gradient. We omitted the latitudinal effect from
ribbon and spotted seal models since, during the spring, these species are most
prevalent near the southern ice edge in the Bering Sea [@conn2014a].

@@

We assessed whether the annual variation in maximum spring sea ice extent in the
Bering Sea influences the seasonal peak of seal haul-out behavior. In
particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
study area with greater than 15% concentration were counted daily to get the total sea
ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily count of grid cells
with greater than 15 percent concentration for each year.

@@

covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "young-of-the-year, subadult, adult male and adult female",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "local solar hour using 6 variables of a Fourier-series basis",
  "Day",   "Continuous", "Bio-logger", 
  "linear, quadratic, and cubic effects of day-of-year",
  "Precip", "Continuous", "NARR", 
  "convective precipitation (kg/m2)",
  "Pressure", "Continuous", "NARR", 
  "atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "air temperatures at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "northerly and easterly vector components for wind converted into a single wind speed via the Euclidean norm",
  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)",
  "Year", "Continuous", "Bio-logger", "For the set of models examining inter-annual variation in sea ice use, we fitted models with the addition of year by day-of-year interactions." 
)

@@

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, we assumed an hourly
Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
the linear predictor was modeled on the logit scale. This is consistent with
previous approaches [@verhoef2014a, @london2012] and only
`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
observations fell between 10% and 90% hourly percent dry.

@@

```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.id = "beardedHOCal", fig.cap="Predicted haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}

@@

The second set of models that included annual variation in haul-out
patterns uncovered significant contributions for linear and quadratic
interactions between day and year in both ribbon seals (day:year,
$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$
= `r ribbon_year_type3$day_year$F.value`; $p$ =
`r ribbon_year_type3$day_year$Prob.F`; day^2^:year,
$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$
= `r ribbon_year_type3$day2_year$F.value`; $p$ =
`r ribbon_year_type3$day2_year$Prob.F`) and spotted seals (day:year,
$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
= `r spotted_year_type3$day_year$F.value`; $p$ =
`r spotted_year_type3$day_year$Prob.F`; day^2^:year,
$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$
= `r spotted_year_type3$day2_year$F.value`; $p$ =
`r spotted_year_type3$day2_year$Prob.F`). Predicted distributions of
haul-out activity were largely unimodal, but varied some among and
within years with respect to the timing and magnitude of haul-out peaks
(Figure \@ref(fig:annualHO)). It is important to note that predicted
variation in annual haul-out patterns likely reflects both process error
and sampling variability. While we did remove any years where only one
deployment in a species + age-sex group was present, there are still
some years where the pattern shown is informed by a small number of
individuals that may not represent population level processes.

```{r annualHO, fig.height = 5.5, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age-sex + year) that included observations from more than one seal are shown. Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}

@@

Predictions of absolute haul-out probability in this paper were somewhat
different than those previously reported for these species, especially for
bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a)] and Conn et al.
[-@conn2014a] used haul-out correction factors with maximums of 0.66 for bearded
seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where maximums
corresponded to times near solar noon in mid-late May. Applying models that
ignore age, sex, and year effects, these probabilities were 0.38, 0.72, and
0.60, respectively, under the current analysis framework. Our current estimates
of haul-out probability reflect increased sample sizes in terms of
number of animals, but also improvements to the way data were prepared prior to
analysis.

We focused this paper on haul-out behavior of bearded, ribbon, and spotted
seals. Ringed seals are also present in the Bering and Chukchi seas but exhibit
a unique complicating factor. Adult ringed seals build subnivean lairs in the
snow on top of the sea ice, where they haul out and pup until snow melt causes
their lairs to collapse [@kelly1990]. Thus, the wet-dry sensor on a bio-logger
could indicate that an animal is hauled out, but if it is within a lair, it is
not available to be detected during an aerial survey. We hope to address
availability of ringed seals using data from satellite tags, replicate survey
tracks, and auxiliary information about snow depth and timing of melt in a
future study.

>>>>

  - \usepackage {newtxtext,newtxmath}

@@

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history
functions, such as pupping, nursing, breeding behavior, and molting [@boveng2009;
@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
complicate predictions about the ultimate effects of changes in sea ice on the
behavior, health, abundance, and distribution of these seals. To date, indices
of seal health sampled during periods of declining sea ice differ regionally
[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
timing of reproductive and molting behavior is generally lacking, so it is
difficult to predict the readiness with which ice-associated seal species will
be able to adapt to future changes (e.g., by adjusting pupping or molting
schedules to earlier dates or different locales). This is further complicated by
the spatio-temporal variation in the phenology of these life history events
within regions and throughout their full ranges. Additionally, trends in
abundance from large scale surveys of these species are unknown, so it is
difficult to assess the effect, if any, declines in sea ice habitat have had, or
will have, on seal densities.

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a resting platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
sea-ice but also within snow lairs. The remoteness of the Bering and Chukchi
seas mean direct scientific observation of seal behavior is impractical. Thus,
bio-logging devices are especially useful tools for collecting key information
on movement and haul-out behavior for these species.

Bio-logging records of time spent out of the water provide valuable data for
identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags and
direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

@@

In this study, we used 16 years of bio-logging data to investigate the haul-out
behavior of bearded, ribbon, and spotted seals in the Bering and Chukchi seas.
Our goals were threefold. First, we wished to establish baseline estimates for
the chronology of haul-out behavior in the critical spring season for each
species across different age and sex classes. Second, we refined estimates of
haul-out availability corrections for aerial surveys in order to improve
estimates of seal abundance. Previously estimated availability correction
factors (e.g., [@bengtson2005a], [@conn2014a], and [@verhoef2014a]) accounted
for variables such as the time-of-day and day-of-year, but did not investigate
meteorological variables that have been shown to influence haul-out behavior of
walruses [@udevitz2009]. Third, we assessed the annual variability
in haul-out timing and possible linkage to changes in the extent of seasonal sea
ice between 2005 and 2020. Our work extends the scope of previous haul-out
analyses, includes the influence of meteorological variability, and investigates
the potential impact of a changing icescape on the behavior of these species.

@@

```{r getdata}
targets::tar_load(analysis_data, store = here::here("_targets"))
targets::tar_load(grid, store = here::here("_targets"))
targets::tar_load(deploy_table, store = here::here("_targets"))
```
Data used in our study came from bio-loggers deployed on bearded, ribbon, and
spotted seals in the Bering, Chukchi, and Beaufort seas by multiple
organizations as part of collaborative investigations from 2005 through 2020.
Seals were captured using nets and bio-loggers attached during ship-based
capture trips and from coastal Alaska communities. Ship-based capture events
occurred during spring near the southern ice edge in the Bering Sea. Land-based
capture events occurred from May to October, generally between the coastal
communities of Scammon Bay, Alaska in the Bering Sea, Utqiaġvik, Alaska in the
Chukchi Sea, and Nuiqsut in the Beaufort Sea from 2005 to 2020. We refer readers
to the primary literature for detailed capture and bio-logger attachment methods
(Supplemental Material, Table).

Haul-out behavior data from bio-loggers deployed on 
`r dplyr::n_distinct(deploy_table$speno)` bearded, ribbon, and spotted seals were
subset to include only records from 1 March to 15 July between 
`r min(lubridate::year(analysis_data$haulout_dt))` and 
`r max(lubridate::year(analysis_data$haulout_dt))`. Bio-loggers were of the
'SPLASH' or 'SPOT' family of tags developed by Wildlife Computers (Redmond,
Washington, USA) and either adhered to the hair on the seal or attached to the
rear flipper inter-digital webbing. The use of bio-loggers adhered to the back
or head provide some benefits over flipper mounted devices (e.g. increased
satellite transmittal rates, locations at sea) but these are lost during the
following annual molt, which, depending on deployment date, limits the duration
of haul-out data they provide. Additionally, bio-loggers attached to the head or
dorsal region are often dry while the seal is at the surface and can slightly
bias the hourly percent-dry values reported by the bio-logger. For this study,
in cases where both bio-logger types were deployed, hourly percent-dry
observations from the flipper tag were preferred.

Field identification of age class can be inexact, particularly when discerning
subadults from adults. Sex and age class (non-dependent _young-of-the-year_,
sexually immature _subadults_, and mature _adults_) were estimated at the time
of deployment by various combinations of length, claw growth ridges
[@mclaren1958a], and pelage characteristics for some species. In the case of
ribbon seals, subadults often have less distinct ribbons than adults. Bearded
seal subadults will often have a spotted pattern in the pelage that is not seen
in adults. Spotted seal pelage cannot be used to reliably discern adults from
subadults. Despite these challenges, we feel the age classifications used in
this analysis are useful in testing for age-related effects on haul-out
behavior. Seals determined to be less than one year were classified as
young-of-the-year. Subadults are those seals likely greater than one year of age
but not yet sexually mature. Adults are likely sexually mature and older than
four years. For those bio-loggers deployed on young-of-the-year and transmitting
into the next year, the age class was advanced to subadult on 1 March of the
following year. Table \@ref(tab:tblDeploy) provides a summary of these
deployments and data received from them.

@@

Haul-out behavior data were recorded and transmitted via the Argos satellite
network as hourly percent-dry timelines. For each hour of a day, the wet/dry
sensor was polled by the tag firmware every few seconds and a percent of the
hour in a dry state was calculated (Figure \@ref(fig:examplePlot)). On board the
bio-logger, hourly percent-dry data were rounded to the nearest 10%
inclusive of 0% and 100% and additional values at 3% and 98%. This
compression resulted in additional data transmission as each message consisted
of two complete 24-hour records. Bio-loggers were, generally, deployed and
programmed in a manner to maximize data transmission during the spring pupping
and molting period; however, hourly percent-dry data were not always
successfully transmitted. This is due to a variety of factors including
satellite coverage, tag availability (i.e. tags mounted to the rear flipper
often do not transmit while at sea), tag performance, duty cycling, and
extra-terrestrial atmospheric interference. Fortunately, missing records do not
seem to greatly bias inference about haul-out probabilities [@conn2012b].

```{r examplePlot, include=TRUE, fig.id="examplePlot", fig.cap="Haul-out behavior observations recorded by a bio-logger deployed on a ribbon seal over two years during the months of March, April, and May. Areas of the plot indicated with an 'X' represent missing data that were not successfully transmitted from the tag."}

@@

p <- p + facet_grid(year~month, scales = "free_x")
p <- p + scale_x_continuous(breaks = c(5,15,25), expand = c(0,0)) +
  scale_y_continuous(breaks = c(4,12,20))
p <- p + theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour="white")) +
  theme(axis.ticks=element_blank()) +
  theme(panel.spacing = unit(0,"line")) +
  ggtitle("Observation records are incomplete because not all data are transmitted",
          subtitle = "percent-dry actogram for a single ribbon seal across two years") +
  xlab("day of month") + ylab("hour (UTC)") 
p <- p + 
  geom_point(aes(x = day, y = hour), data = missing_hours, 
             color = "black", shape = 4, size=0.8)
p
```

Of key interest in this study was the relationship between haul-out behavior and
weather covariates that vary with time and animal location. We explored the use
of a continuous-time correlated random walk [@johnson2008a] movement model
to predict locations at specific times. However, the sparse nature of data from
some bio-loggers, especially those mounted to the rear flipper, resulted in poor
modeling performance or convergence issues. For this study, we calculated a
weighted average daily location where the inverse of the estimated Argos or
FastLoc GPS location error was used for the weight. Each Argos location estimate
was assigned an error radius based on either the categorical location quality (
_3_=250m, _2_=500m, _1_=1500m, _0_=2500m [@lopez2013]; we chose 2500m for
location classes _A_ and _B_) or, when available, the estimated error radius
from the Argos Kalman filter algorithm. Location estimates from FastLoc GPS were
all assigned an error radius of 50m. Any days where haul-out observations were
present but location data were missing we used the last calculated weighted
average daily location, and any days where the location intersected with land
were removed from the data set. We recognized that bearded and spotted seals
haul out on land. However, assessing the relationship between haul-out behavior
and weather covariates and their availability for aerial surveys on land was
outside the scope of this study. Additionally, any daily locations on land are
likely more reflective of coordinate averaging and measurement error instead of
actual use of coastal features. Figure \@ref(fig:dataMap) shows the spatial
distribution weighted locations with available haul-out behavior data for each
species across the study area.

@@

Models for all species include the following effects: day-of-year, solar hour,
temperature, wind speed, barometric pressure, precipitation, wind chill
(represented by a *wind:temperature* interaction [@udevitz2009]), and
day-of-year and time-of-day (solar hour) interactions to permit diurnal patterns
to change from March to July. Ribbon and spotted seal models included age-sex
class and interactions between day-of-year and age-sex class, but we omitted
these from bearded seal models due to poor representation of age-sex classes
(Table \@ref(tab:tblDeploy)). Bearded seal models included a latitudinal effect
(and an interaction with day-of-year), since bearded seals occupy a substantial
range and we were interested in possible differences in the timing of haul-out
behavior along a latitudinal gradient. We omitted the latitudinal effect from
ribbon and spotted seal models because, during the spring, these species are most
prevalent near the southern ice edge in the Bering Sea [@conn2014a].

@@

We assessed whether the annual variation in maximum spring sea ice extent in the
Bering Sea influences the seasonal peak of seal haul-out behavior. In
particular, we used sea ice concentration data from the Nimbus-7 SMMR and DMSP
SSM/I-SSMIS Passive Microwave Dataset, Version 1 [@cavalieri1996] to calculate
maximum sea ice extent. All sea ice concentration grid cells (25 km^2^) in the
study area with greater than 15% concentration were counted daily to get the total sea
ice extent for each day between 15 February and 15 July across all years.
Maximum spring sea ice extent was simply the largest daily count of grid cells
with greater than 15% concentration for each year.

@@

covariate_tbl <- tibble::tribble(
  ~Covariate, ~Type, ~Source, ~Description,
  "Age-sex class",   "Categorical", "Field Assessment", 
  "young-of-the-year, subadult, adult male and adult female",
  "Hour", "Continuous; Fourier basis", "Bio-logger", 
  "local solar hour using 6 variables of a Fourier-series basis",
  "Day",   "Continuous", "Bio-logger", 
  "linear, quadratic, and cubic effects of day-of-year",
  "Precip", "Continuous", "NARR", 
  "convective precipitation (kg/m^2^)",
  "Pressure", "Continuous", "NARR", 
  "atmospheric pressure at sea level (kPa)",
  "Temp", "Continuous", "NARR",
  "air temperatures at 2m above the earth’s surface",
  "Wind", "Continuous", "NARR", 
  "northerly and easterly vector components for wind converted into a single wind speed via the Euclidean norm",
  "Northing", "Continuous", "Bio-logger","latitude divided by the mean latitude across all locations (for bearded seals only)",
  "Year", "Continuous", "Bio-logger", "For the set of models examining inter-annual variation in sea ice use, we fitted models with the addition of year by day-of-year interactions." 
)

@@

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the glmmLDTS package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, we assumed an hourly
Bernoulli response (i.e., whether tags were mostly dry or mostly wet) where
the linear predictor was modeled on the logit scale. This is consistent with
previous approaches [@verhoef2014a, @london2012] and only
`r (1 - (portion_10(analysis_data) + portion_90(analysis_data))) * 100`% of our
observations fell between 10% and 90% hourly percent-dry.

@@

```{r beardedHOCal, fig.height = 4.5, include=TRUE, fig.cap="Predicted haul-out probability of bearded seals (all ages and sex classes) from 1 March to 15 July for all and sex classes."}

@@

The second set of models that included annual variation in haul-out
patterns uncovered significant contributions for linear and quadratic
interactions between day and year in both ribbon seals (day:year,
$F_{`r paste(ribbon_year_type3$day_year$Num.df, ribbon_year_type3$day_year$Den.df, sep=",")`}$
= `r ribbon_year_type3$day_year$F.value`; $p$ =
`r ribbon_year_type3$day_year$Prob.F`; day^2^:year,
$F_{`r paste(ribbon_year_type3$day2_year$Num.df, ribbon_year_type3$day2_year$Den.df, sep=",")`}$
= `r ribbon_year_type3$day2_year$F.value`; $p$ =
`r ribbon_year_type3$day2_year$Prob.F`) and spotted seals (day:year,
$F_{`r paste(spotted_year_type3$day_year$Num.df, spotted_year_type3$day_year$Den.df, sep=",")`}$
= `r spotted_year_type3$day_year$F.value`; $p$ =
`r spotted_year_type3$day_year$Prob.F`; day^2^:year,
$F_{`r paste(spotted_year_type3$day2_year$Num.df, spotted_year_type3$day2_year$Den.df, sep=",")`}$
= `r spotted_year_type3$day2_year$F.value`; $p$ =
`r spotted_year_type3$day2_year$Prob.F`). Predicted distributions of
haul-out activity were largely unimodal, but varied some among and
within years with respect to the timing and magnitude of haul-out peaks
(Figure \@ref(fig:annualHO)). It is important to note that predicted
variation in annual haul-out patterns likely reflects both process error
and sampling variability. While we did remove any years where only one
deployment in a species + age-sex group was present, there are still
some years where the pattern shown is informed by a small number of
individuals that may not represent population level patterns.

```{r annualHO, fig.height = 5.5, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age-sex + year) that included observations from more than one seal are shown. Additionally, any groups where data were only available after 1 June or before 1 May are not included."}

@@

Predictions of absolute haul-out probability in this paper were somewhat
different than those previously reported for these species, especially for
bearded seals. For instance, Ver Hoef et al. [-@verhoef2014a] and Conn et al.
[-@conn2014a] used haul-out correction factors with maximums of 0.66 for bearded
seals, 0.62 for ribbon seals, and 0.54 for spotted seals, where maximums
corresponded to times near solar noon in mid-late May. Applying models that
ignore age, sex, and year effects, these probabilities were 0.38, 0.72, and
0.60, respectively, under the current analysis framework. Our current estimates
of haul-out probability reflect increased sample sizes in terms of
number of animals, but also improvements to the way data were prepared prior to
analysis.

We focused this paper on haul-out behavior of bearded, ribbon, and spotted
seals. Ringed seals are also present in the Bering and Chukchi seas but exhibit
a unique complicating factor. Adult ringed seals build subnivean lairs under the
snow on top of the sea ice, where they haul out and where females rear pups until
conditions are good for basking [@frost2004]. Thus, the wet-dry sensor on a bio-logger
could indicate that an animal is hauled out, but if it is within a lair, it is
not available to be detected during an aerial survey. We hope to address
availability of ringed seals using data from satellite tags, replicate survey
tracks, and auxiliary information about snow depth and timing of melt in a
future study.
